---
title: "Seeding Report Card"
author: Thomas Buehrens, WDFW (tbuehrens@dfw.wa.gov),
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>
```

------------------------------------------------------------------------

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

------------------------------------------------------------------------

```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```


## Functions
Here, we define some functions to fit two forms of Ricker models and a Beverton Holt Model
```{r set_up, echo = TRUE, message = FALSE}
#function to install or load packages
install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg))
    install.packages(create.pkg, dependencies = TRUE,repos = "http://cran.us.r-project.org")
  sapply(pack, require, character.only = TRUE)
}

#logit transform
logit<-function(x){log(x/(1-x))}

#inverse logit transform
ilogit<-function(x){exp(x)/(1+exp(x))}

#fit ricker model with separate hatchery and wild productivity
nll_ricker<-function(par,P,R,S){
  logalpha<-par[1]
  alpha=exp(logalpha)
  loginvbeta<-par[2]
  beta<-1/exp(loginvbeta)
  logitb<-par[3]
  b<-ilogit(logitb)
  logsigma<-par[4]
  sigma<-exp(logsigma)
  R_hat<-(alpha * (1-P) * S + alpha * b * P * S) * exp(-1 * beta * S) 
  resid<-log(R_hat)-log(R)
  nll=sum(-1*log(dnorm(resid,mean = 0,sd = sigma)))
}

#fit linearized ricker model with covariate effect on pHOS
nll_ricker2<-function(par,P,R,S){
  logalpha<-par[1]
  alpha=exp(logalpha)
  loginvbeta<-par[2]
  beta<-1/exp(loginvbeta)
  b<-par[3]
  logsigma<-par[4]
  sigma<-exp(logsigma)
  logRS_hat<- log(alpha) - beta * S + b * P 
  resid<-logRS_hat-log(R/S)
  nll=sum(-1*log(dnorm(resid,mean = 0,sd = sigma)))
}

#fit beverton-holt model with separate hatchery and wild productivity
nll_bh<-function(par,P,R,S){
  logalpha<-par[1]
  alpha=exp(logalpha)
  logbeta<-par[2]
  beta<-exp(logbeta)
  logitb<-par[3]
  b<-ilogit(logitb)
  logsigma<-par[4]
  sigma<-exp(logsigma)
  R_hat<-(alpha * (1-P) * S + alpha * b * P * S)/(1 + (alpha * (1-P) * S + alpha * b * P * S) /beta) 
  resid<-log(R_hat)-log(R)
  nll=sum(-1*log(dnorm(resid,mean = 0,sd = sigma)))
}

#helper function for transforming ricker model params to link scales for estimation (to constrain them to proper domains; e.g., alpha to be positive)
ricker_transform<-function(real_par){
  return(c(
    logalpha=log(real_par[1]),
    loginvbeta=log(1/real_par[2]),
    logitb=logit(real_par[3]),
    logsigma=log(real_par[4])
    )
  )
}

#ricker back transform
ricker_inv_transform<-function(par){
  par<-as.vector(par)
  return(c(
    alpha=exp(par[1]),
    beta=1/exp(par[2]),
    b=ilogit(par[3]),
    sigma=exp(par[4])
    )
  )
}

#helper function for transforming ricker model version 2 params to link scales for estimation (to constrain them to proper domains; e.g., alpha to be positive)
ricker_2_transform<-function(real_par){
  return(c(
    logalpha=log(real_par[1]),
    loginvbeta=log(1/real_par[2]),
    b=real_par[3],
    logsigma=log(real_par[4])
  )
  )
}

#ricker back transform
ricker_2_inv_transform<-function(par){
  par<-as.vector(par)
  return(c(
    alpha=exp(par[1]),
    beta=1/exp(par[2]),
    b=par[3],
    sigma=exp(par[4])
  )
  )
}


bh_transform<-function(real_par){
  return(c(
    logalpha=log(real_par[1]),
    logbeta=log(real_par[2]),
    logitb=logit(real_par[3]),
    logsigma=log(real_par[4])
  )
  )
}

#beverton-holt back transform
bh_inv_transform<-function(par){
  par<-as.vector(par)
  return(c(
    alpha=exp(par[1]),
    beta=exp(par[2]),
    b=ilogit(par[3]),
    sigma=exp(par[4])
  )
  )
}
```

## Load Packages
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
install_or_load_pack(pack = c("tidyverse","RColorBrewer","kableExtra","gridExtra","viridis"))
```

## Load Data
```{r user_inputs_data, message = FALSE, warning = FALSE,results = "show",fig.align='center',fig.show=TRUE, fig.cap="Figure 1. Stock recruit data"}
dat<-read_csv("Tilton_CK_juv_recruits_vs_female_spawners.csv")

dat%>%
  kbl(caption = paste0("Table 1. Example data to fit SR models to develop a seeding report card (in this case Tilton Chinook female hatchery and wild Spawners and the smolt recruits they produced)"),digits =2)%>%
  kable_classic(full_width = F, html_font = "Cambria")


dat%<>%
  #filter(`Brood Year`>2010)%>%
  mutate(S=Wild+Hatchery,
         P=Hatchery/S,
         R=Recruits,
         lnrs=log(R/S)
         )%>%
  dplyr::rename(brood_year=`Brood Year`)%>%
  filter(!is.na(P),R>0)

ggplot(dat,aes(x=S,y=R,color=P))+
  geom_point(size=2)+
  xlim(0,NA)+
  ylim(0,NA)+
  labs(x="Spawners",y="Recruits",color="pHOS")+
  scale_color_viridis_c()
```


# Fit SR Models
NOTE: Make sure to check that the convergence value is 0 for each model (i.e., they each converged...otherwise need to change initial values)

Here we will fit three SR models:

First, a modified Ricker model where the total spawners $S$ are multiplied by pHOS $p$ and pNOS $(1-p)$ to derive hatchery and wild spawners and allow us to model their productivities discretely.The recruitment $R$ is function of wild and hatchery spawners where the productivity $\alpha$ for hatchery spawners is multiplied by the relative reproductive success $b$. Density dependence, which is a function of $\beta$ (the inverse of the number of spawners generating the maximum recruitment $Rmax$), is related to total spawner abundance independent of origin:
1) $$R = (\alpha (1-p) S + apbS) e^{-\beta S}  e^\epsilon$$
where parameters were fit subject to the following constraints:
$$0<\alpha,\sigma, \beta$$
$$0<b<1$$
$$\epsilon \sim normal(0,\sigma)$$

Second, a Ricker model where recruitment $R$ is modeled conventionally as a function of total spawners, but a multiplicative covariate effect $b$ on pHOS $p$ is added to the right hand side of the equation:
2) $$R = \alpha S e^{-\beta S} e^{bp} e^\epsilon$$
where parameters were fit subject to the following constraints:
$$0<\alpha,\sigma, \beta$$
$$\epsilon \sim normal(0,\sigma)$$

Third, a Beverton-Holt model where recruitment $R$ is modeled as a function of hatchery spawners $pS$ and wild spawners $(1-p)S$ and the productivity of the hatchery spawners is multiplied by their estimated relative reproductive success $b$. In this case there is one maximum recruitment value for the population $\beta$ also known as $Rmax$ but different productivity depending on origin.
3) $$R = \frac {\alpha (1-p) S + \alpha p b S}{1 + {\frac {\alpha (1-p) S + \alpha p b S} \beta} }e^\epsilon $$
where parameters were fit subject to the following constraints:
$$0<\alpha,\sigma, \beta$$
$$0<b<1$$
$$\epsilon \sim normal(0,\sigma)$$


```{r Analysis_v2, message=FALSE, warning=FALSE, results="hide",cache=TRUE}
alpha_guess<-max(exp(dat$lnrs)) #guess alpha parameter
ricker_beta_guess<--lm(lnrs~S,data=dat)$coef[2] #guess beta parameter
bh_beta_guess<-max(dat$Recruits) #guess beta parameter
b_guess<-0.1 #guess Relative Reproductive Success of Hatchery Fish
sigma_guess<-0.4 #guess log-normal residual error sd of SR model


result <- optim(par = ricker_transform(real_par=c(alpha_guess, ricker_beta_guess,b_guess,sigma_guess)), fn = nll_ricker, R = dat$R , S = dat$S,P=dat$P) # par is the initial values of the coefficients
result$par<-ricker_inv_transform(par=result$par) # the maximum likelihood estimates of the coefficients
result$value # the minimum value of the negative log-likelihood
result$convergence # the convergence code (0 means successful convergence)

result2 <- optim(par = ricker_2_transform(real_par=c(alpha_guess, ricker_beta_guess,-(0.2/b_guess),sigma_guess)), fn = nll_ricker2,  R = dat$R , S = dat$S,P=dat$P) # par is the initial values of the coefficients
result2$par<-ricker_2_inv_transform(par=result2$par) # the maximum likelihood estimates of the coefficients
result2$value # the minimum value of the negative log-likelihood
result2$convergence # the convergence code (0 means successful convergence)

result3 <- optim(bh_transform(real_par= c(alpha_guess, bh_beta_guess,b_guess,sigma_guess)), fn = nll_bh,  R = dat$R , S = dat$S,P=dat$P) # par is the initial values of the coefficients
result3$par<-bh_inv_transform(par=result3$par) # the maximum likelihood estimates of the coefficients
result3$value # the minimum value of the negative log-likelihood
result3$convergence # the convergence code (0 means successful convergence)
```


# Results
```{r Results, message=FALSE, warning=FALSE, results="show",cache=TRUE}

params<-tibble(model = c("Ricker", "Ricker version 2", "Beverton-Holt"))%>%
                 bind_cols(bind_rows(result$par,result2$par,result3$par))%>%
                 set_names(c("model","alpha","beta","b","sigma")
                 )%>%
  mutate(Rmax= ifelse(model=="Beverton-Holt",beta,alpha*(1/beta)*exp(-beta*(1/beta))))

params%>%
  kbl(caption = paste0("Table 2. SR params from three models; alpha is max R/S for all models; beta is 1/spawners that produce max recruits for ricker but is equal to max recruits for Beverton-Holt; b is relative reproductive success of hatchery fish for Ricker version 1 and for Beverton-Holt but is the slope of a coefficient for the effect of pHOS on recruitment for Ricker version 2; sigma is log-normal residual error sd; Rmax is maximum recruitment."),digits =4)%>%
  kable_classic(full_width = F, html_font = "Cambria")


  
fits<-expand_grid(S=seq(1,5000,length.out=100),P=seq(0,0.99,length.out=100))%>%
  mutate(
    Ricker = (result$par[1] * (1-P) * S + result$par[1] * result$par[3] * P * S) * exp(-1 * result$par[2] * S),
    `Ricker version 2` = exp(log(result2$par[1]) - result2$par[2] * S + result2$par[3] * P)*S,#log(P/(1-P))
    `Beverton-Holt` =(result3$par[1] * (1-P) * S + result3$par[1] * result3$par[3] * P * S)/(1 + (result3$par[1] * (1-P) * S + result3$par[1] * result3$par[3] *P * S) /result3$par[2])
  )


fits2<-expand_grid(S=seq(1,5000,length.out=100),P=seq(0,1,length.out=11))%>%
  mutate(
    Ricker = (result$par[1] * (1-P) * S + result$par[1] * result$par[3] * P * S) * exp(-1 * result$par[2] * S),
    `Ricker version 2` = exp(log(result2$par[1]) - result2$par[2] * S + result2$par[3] * P)*S,#log(P/(1-P))
    `Beverton-Holt` =(result3$par[1] * (1-P) * S + result3$par[1] * result3$par[3] * P * S)/(1 + (result3$par[1] * (1-P) * S + result3$par[1] * result3$par[3] *P * S) /result3$par[2])
  )


#color_palette <- brewer.pal(11,"RdBu")

p1<-ggplot(fits,aes(x=S,y=P),alpha=0.5)+
  geom_tile(aes(fill=as.factor(round(Ricker/max(params$Rmax[params$model=="Ricker"]),1))))+
  scale_fill_viridis_d(name = "R/Rmax", guide=guide_legend(override.aes = list(shape=NA)))+
  # scale_fill_manual(values = alpha(color_palette,0.35), name = "R/Rmax",
  #                   guide=guide_legend(override.aes = list(shape=NA)))+
  geom_label(dat,mapping=aes(x=S,y=P),label=dat$brood_year,color="black",fill=NA,label.size=NA,size=3)+
  xlim(0,NA)+
  ylim(0,NA)+
  ylab("pHOS")+
  xlab("Female Spawners")+
  labs(title="Ricker version 1")+
  theme(aspect.ratio = 1)#+
  #theme(legend.position = "none")

p2<-ggplot(fits,aes(x=S,y=P),alpha=0.5)+
  geom_tile(aes(fill=as.factor(round(`Ricker version 2`/max(params$Rmax[params$model=="Ricker version 2"]),1))))+
  scale_fill_viridis_d(name = "R/Rmax", guide=guide_legend(override.aes = list(shape=NA)))+
  # scale_fill_manual(values = alpha(color_palette,0.35), name = "R/Rmax",
  #                   guide=guide_legend(override.aes = list(shape=NA)))+
  geom_label(dat,mapping=aes(x=S,y=P),label=dat$brood_year,color="black",fill=NA,label.size=NA,size=3)+
  xlim(0,NA)+
  ylim(0,NA)+
  ylab("pHOS")+
  xlab("Female Spawners")+
  labs(title="Ricker version 2")+
  theme(aspect.ratio = 1)#+
  #theme(legend.position = "none")


p3<-ggplot(fits,aes(x=S,y=P),alpha=0.5)+
  geom_tile(aes(fill=as.factor(round(`Beverton-Holt`/max(params$Rmax[params$model=="Beverton-Holt"]),1))))+
  scale_fill_viridis_d(name = "R/Rmax", guide=guide_legend(override.aes = list(shape=NA)))+
  # scale_fill_manual(values = alpha(color_palette,0.35), name = "R/Rmax",
  #                   guide=guide_legend(override.aes = list(shape=NA)))+
  geom_label(dat,mapping=aes(x=S,y=P),label=dat$brood_year,color="black",fill=NA,label.size=NA,size=3)+
  xlim(0,NA)+
  ylim(0,NA)+
  ylab("pHOS")+
  xlab("Female Spawners")+
  labs(title="Beverton-Holt")+
  theme(aspect.ratio = 1)#+
  #theme(legend.position = "right")

figure3<-ggplot(
  fits2%>%
    pivot_longer(cols=!c(S,P),names_to = "model",values_to = "R")%>%
    arrange(model,S,P)%>%
    group_by(S,R,P,model)%>%
    left_join(params%>%
                dplyr::select(model,Rmax)
              )%>%
    arrange(model,rev(P),S)
    
  ,aes(x=S,y=R,group=interaction(model,P)))+
  geom_line(aes(color=as.factor(P)),size=2)+
  facet_wrap(~model, nrow = 1)+
  xlim(0,NA)+
  ylim(0,NA)+
  labs(x="Spawners",y="Recruits",color="pHOS")+
  scale_color_viridis_d()


figure4<-ggplot(
  fits2%>%
    pivot_longer(cols=!c(S,P),names_to = "model",values_to = "R")%>%
    arrange(model,S,P)%>%
    group_by(S,R,P,model)%>%
    left_join(params%>%
                dplyr::select(model,Rmax)
              )%>%
    arrange(model,rev(P),S)
    
  ,aes(x=S,y=R/Rmax,group=interaction(model,P)))+
  geom_line(aes(color=as.factor(P)),size=2)+
  facet_wrap(~model, nrow = 1)+
  xlim(0,NA)+
  ylim(0,NA)+
  labs(x="Spawners",y="R/Rmax",color="pHOS")+
  scale_color_viridis_d()

```

```{r message=FALSE, warning=FALSE, results="hide", fig.align='center',fig.show=TRUE, fig.cap="Figure 2.The proportion of full seeding predicted by three stock recruit models as a function of spawner abundance and pHOS. Data points from each year are shown on the graph, depicting the predicted (as opposed to observed) seeding for that year."}
#grid.arrange(p1, p2, p3, ncol = 1, respect=T)
p1
p2
p3
```

```{r message=FALSE, warning=FALSE, results="hide", fig.align='center',fig.show=TRUE, fig.cap="Figure 3.The effect of pHOS on the expected recruitment relationship based on three SR fits."}
figure3
```


```{r message=FALSE, warning=FALSE, results="hide", fig.align='center',fig.show=TRUE, fig.cap="Figure 4.The effect of pHOS on the expected % of full seeding (R/Rmax) depending on spawner abundance and pHOS."}
figure4
```


# Possible Use Cases
This analysis may be useful for:
1) empirical estimates of productivity (alpha), capacity (Rmax), and relative reproductive success (b for Ricker version 1 and Beverton-Holt models)
2) to determine what level of seeding is accomplished based on various spawner abundances and pHOS levels (for in-season and post-season evaluations)
3) to ground truth heuristic models like AHA 
